Methods for Analyzing Models of Rod-Shaped Bacteria

### 2025-07-27 paper/main.tex
switch subsections
remove some TODOs

%###################################################################################################
\section{Introduction (TODO)}
\label{section:introduction}

@@

%###################################################################################################
\section{Overview (TODO)}
This work consists of many subcomponents which can be combined in order to achieve various tasks.
Figure~\ref{fig:flowchart-project-structure} shows a high-level overview of all components.
We assume that bacterial rods can be represented by a colleciton of vertices $\vec{v}_i$.
This assumption alone is enough to enable a cascade of other components.
Using the spatial representation, we provide visualization methods that create 3D models from the
given vertices which can then either be rendered or projected to a 2D plane.
The data extraction component determines the vertices $\vec{v}_i$ from given cell masks which
correspond to microscopic images.
This allows us in particular to turn any sequence of images (movie) into a sequence of bacterial
positions.
The dynamics of the vertices $\vec{v}_i$ are determined by a computational model which is specific
to the respective use-case and should be defined by the researcher.
It merely needs to utilize the aforementioned spatial representation of bacterial rods as a
collection of vertices $\vec{v}_i$.
This work will provide one example of how such a model can be defined.
Notably, the visualization and data extraction components do not require any knowledge about the
computational model, thus allowing us to reuse their functionality across multiple models.

@@

% TODO write more text here
% Use some parts of the proposal text

%###################################################################################################
\section{Methodology}
% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation of Rod-Shaped Bacteria}
% TODO CITATIONS
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{v}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{v}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{v}_{i-1},\vec{v}_i$ can be used when modeling the
dynamics of these bacteria.

>>>>

%###################################################################################################
\section{Introduction}
\label{section:introduction}

@@

%###################################################################################################
\section{Overview}
This work consists of many subcomponents which can be combined in order to achieve various tasks.
Figure~\ref{fig:flowchart-project-structure} shows a high-level overview of all components.
We assume that bacterial rods can be represented by a colleciton of vertices $\vec{v}_i$.
This assumption alone is enough to enable a cascade of other components.
Using the spatial representation, we provide visualization methods that create 3D models from the
given vertices which can then either be rendered or projected to a 2D plane.
The data extraction component determines the vertices $\vec{v}_i$ from given cell masks which
correspond to microscopic images.
This allows us in particular to turn any sequence of images (movie) into a sequence of bacterial
positions.
The dynamics of the vertices $\vec{v}_i$ are determined by a computational model which is specific
to the respective use-case and should be defined by the researcher.
It merely needs to utilize the aforementioned spatial representation of bacterial rods as a
collection of vertices $\vec{v}_i$.
This work will provide one example of how such a model can be defined.
Notably, the visualization and data extraction components do not require any knowledge about the
computational model, thus allowing us to reuse their functionality across multiple models.

@@

%###################################################################################################
\section{Methodology}
% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation of Rod-Shaped Bacteria}
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{v}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{v}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{v}_{i-1},\vec{v}_i$ can be used when modeling the
dynamics of these bacteria.
